Online Verification of Commutativity

### 2020-07-14 acmPaper/paper.tex
Transcribe algorithms

@@

\begin{abstract}
Commutative diagrams arise in many systems, such as an implicit type conversion graph in some programming languages.
It can be useful in program analysis to verify that a diagram representing conversions remains commutative over the course of its construction.
A naive approach to solving the problem would take factorial time, and the presence of cycles could potentially lead to an infinite number of paths to check. Previous work provides an algorithm for the special case of acyclic diagrams that verifies that a given diagram commutes in $O(|V|^4|E|^2)$ time, where V is the set of vertices in the diagram, and E, the set of edges. 
We present an algorithm that given a diagram and an edge to add to it, verifies that the resultant diagram commutes in $O(|V|(|E| + |V|)$ time. For the case when checking the equality of paths is expensive, we present an optimization that runs in $O(V^4)$ time but reduces to the minimum possible number of equality checks. We evaluate the algorithms against baselines, and see its practical application in the compiler of a domain specific language for geometry types. As a demonstration of the more general applicability of the algorithm, we use it to identify discrepancies in a  currency conversion graph.
\end{abstract}
%
%
%

@@

We start with the set of all parallel pairs in the diagram (with the new edge added in), and pare it down to be finite by handling cycles.
We find all simple cycles in the diagram, using a procedure like Johnson's algorithm. 
\xxx[AK]{TODO cite the paper.}
We verify for each cycle that a single traversal starting on any node is equal to the identity function by adding pairs of (cycle starting at node, identity function) for each node in the function to the output verification set.
\xxx[AK]{Need cleaner way to put this.}
As cycles must now be the identity, for any pair in the set of all parallel paths, any instance of a cycle can be removed to obtain an equivalent pair with shorter, cycle free paths.
If the shorter pair has equal paths then the paths in the original pair must also be equal to each other.
It is therefore safe to remove all pairs of paths with cycles, leaving only parallel pairs where neither path has a cycle.
Since there are only a finite number of paths in the graph with no cycles, (to the order of n! as a path is an ordering on nodes), and the parallel pairs are a subset of the possibilities generated by choosing a first and second member from this finite set, the resultant set of all parallel pairs must be finite.
\xxx[AK]{TODO Check if O(n!n) = O(n). What about $O((n!)^2)$?}
The output of the algorithm, then, is the union of the set of cycle verification pairs, and the parallel pairs found by doing an all-acyclic-path search from every node in the diagram with the new edge added.

@@

To present this idea diagrammatically, we represent path phases with arrows (note that these are the composition of many edges, not a single edge). The new edge is represented with a dashed arrow.

\begin{center}
\begin{tikzpicture}[scale=0.2]
\tikzstyle{every node}+=[inner sep=0pt]
\draw [black] (18.1,-36.1) circle (3);
\draw (18.1,-36.1) node {$S$};
\draw [black] (50.2,-36.1) circle (3);
\draw (50.2,-36.1) node {$T$};
\draw [black] (26,-22.9) circle (3);
\draw (26,-22.9) node {$F_1$};
\draw [black] (41.3,-22.9) circle (3);
\draw (41.3,-22.9) node {$F_2$};
\draw [dashed] (21.1,-36.1) -- (47.2,-36.1);
\fill [black] (47.2,-36.1) -- (46.4,-35.6) -- (46.4,-36.6);
\draw (34.15,-35.6) node [above] {new edge};
\draw [black] (24.46,-25.47) -- (19.64,-33.53);
\fill [black] (19.64,-33.53) -- (20.48,-33.1) -- (19.62,-32.58);
\draw [black] (29,-22.9) -- (38.3,-22.9);
\fill [black] (38.3,-22.9) -- (37.5,-22.4) -- (37.5,-23.4);
\draw [black] (48.52,-33.61) -- (42.98,-25.39);
\fill [black] (42.98,-25.39) -- (43.01,-26.33) -- (43.84,-25.77);
\end{tikzpicture}
\end{center}
Here, (S, $F_1$, $F_2$, T) is a two flip tolerant path. ($F_1$, S, T, $F_2$) is a new path, created because of the addition of (S,T), that conflicts with ($F_1$, $F_2$).

@@

A diagram's structure can result in some redundancy, in the sense that verifying a subset of pairs can imply the verification of path pairs that aren't in the subset.
For example, in the figure below, verifying $p_1$ and $p_2$ implies that the paths in $p_3$ are equal too.

@@

$p_1$=($f_1;f_2$, $g_1$)\\
$p_2$=($j_1;j_2$, $h_1$)\\
$p_3$=($f_1;h_1$, $f_1;j_2$)
\end{center}

\xxx[dg]{Should be a figure}
The approach in this algorithm, at a high level, is to define a function that takes in a subset of pairs and returns the subset of pairs whose verification is implied by verifying the input set. Then, greedy elimination of redundancies is performed until a minimal set is reached.

@@

A complete implementation of the described algorithm is included in the appendix.

\xxx[dg]{Probably put this in the appendix}
\xxx[dg]{Give a summary of the algorithm though, breaking down important parts.  I think that's mostly been done above, but some actual code snippets might help}
\begin{verbatim}
Graph existingGraph;
R_s = {}
for each node v in V:
    # Let the subsection of the graph reachable from v be S.
    S = existingGraph.extractReachableSection(v)
    # Create a minimum spanning tree of S
    T = createMinimumSpanningTree(S)
    # Find the excluded edges
    excludedEdges = S.edges - T.edges
    # Use these edges to create bilinkings
    for each edge e in excludedEdges:
        firstPath = T.findPath(source: e.source, sink: e.sink)
        R_s.addElement(new Bilinking(firstPath, e)
return R_s
\end{verbatim}

\begin{verbatim}
Find a spanning set Rs, = [r_1, ... ,r_k].

R := Rs,
    for i:=l to K do
        if r_i in sigma(R - r_i) then R:=R - r_i;
return R.

sigma(inputSet S, codomain = Rs):
    output = {}
    for bilinking in Rs:
        smallerPairs = {}
        for candidateBilinking in Rs:
            if exists path(source: bilinking.source, sink:candidateBilinking.source) and
            exists path(source: candidateBilinking.sink, sink: bilinking.sink):
                smallerPairs.add(candidateBilinking)
        # Check if there's a linear dependence between the bilinking and smallerPairs
        # vectorize(pair) returns a vector with a dimension for each edge in the graph,
        # with a 1 if pair includes that edge and 0 if it doesn't.
        matrix = [vectorize(pair) for pair in smallerPair]
        matrix = matrix+[vectorize(bilinking)]
        # if there is a linear dependence
        if determinant(matrix)%2==0:
            output.add(bilinking)
    return output

\end{verbatim}

@@

We present a polynomial time solution to the \textsc{verification set} problem.
Like with the baseline, we do not concern ourselves with parallel pairs where neither or both paths pass through the new edge.
The key observation that allows us to improve on the online baseline is that, for a given source and sink pair, only a single parallel pair needs to be verified. 
This is an implication of Theorem \ref{reductionRule}, expanded on later.
Theorem \ref{verifyingSet} shows that should our selected set of pairs along with cycles passing through the new edge be verified commutative, the entire diagram must commute. The approach is to identify a parallel pair with exactly one path through the edge for each (source, sink) pair.

\begin{verbatim}
Graph existingGraph;
Edge newEdge;

Set parallelPairs = new Set();
for S in existingGraph.Nodes():
    for T in existingGraph.Nodes():
        try:
            Path pathWithNewEdge = 
                FindPath(
                    graph: existingGraph, 
                    sourceNode: S,
                    sinkNode: newEdge.Source()) +
                newEdge as Path +
                FindPath(
                    graph: existingGraph, 
                    sourceNode: newEdge.Sink(), 
                    sinkNode: T)
            if (S==T):
                // Cycles are a speacial case.
                pathInOldGraph = identity(S)
            else:
                Path pathInOldGraph = FindPath(
                    graph: existingGraph, 
                    sourceNode: S, 
                    sinkNode: T);
            parallelPairs.add((pathInOldGraph, pathWithNewEdge));
        catch (pathFindingFailed Exception):
            // No comparable pairs from node {S} to node {T} 
            // that need to be checked
            continue

output parallelPairs;
\end{verbatim}
\begin{algorithm}
    \caption{Polynomial time algorithm}
    \label{polynomial}
    \KwData{placeholder}
    \KwResult{placeholder}
\end{algorithm}

@@

Consider the following situation:

\begin{center}
\begin{tikzpicture}[scale=0.2]
\tikzstyle{every node}+=[inner sep=0pt]
\draw [black] (16.6,-34.1) circle (3);
\draw (16.6,-34.1) node {$S$};
\draw [black] (56.8,-33.6) circle (3);
\draw (56.8,-33.6) node {$T$};
\draw [black] (28,-20.1) circle (3);
\draw (28,-20.1) node {$P_1$};
\draw [black] (44.5,-20.1) circle (3);
\draw (44.5,-20.1) node {$P_2$};
\draw [black] (28,-44.9) circle (3);
\draw (28,-44.9) node {$Q_1$};
\draw [black] (44.5,-44.9) circle (3);
\draw (44.5,-44.9) node {$Q_2$};
\draw [black] (31,-20.1) -- (41.5,-20.1);
\fill [black] (41.5,-20.1) -- (40.7,-19.6) -- (40.7,-20.6);
\draw (36.25,-20.6) node [below] {$g_1$};
\draw [black] (26.11,-22.43) -- (18.49,-31.77);
\fill [black] (18.49,-31.77) -- (19.39,-31.47) -- (18.61,-30.84);
\draw (21.74,-25.67) node [left] {$f_1$};
\draw [black] (25.82,-42.84) -- (18.78,-36.16);
\fill [black] (18.78,-36.16) -- (19.01,-37.08) -- (19.7,-36.35);
\draw (23.57,-39.02) node [above] {$f_2$};
\draw [black] (31,-44.9) -- (41.5,-44.9);
\fill [black] (41.5,-44.9) -- (40.7,-44.4) -- (40.7,-45.4);
\draw (36.25,-45.4) node [below] {$g_2$};
\draw [black] (54.59,-35.63) -- (46.71,-42.87);
\fill [black] (46.71,-42.87) -- (47.64,-42.7) -- (46.96,-41.96);
\draw (49.21,-38.76) node [above] {$h_2$};
\draw [black] (54.78,-31.38) -- (46.52,-22.32);
\fill [black] (46.52,-22.32) -- (46.69,-23.25) -- (47.43,-22.57);
\draw (51.19,-25.39) node [right] {$h_1$};
\draw [black] (28,-23.1) -- (28,-41.9);
\fill [black] (28,-41.9) -- (28.5,-41.1) -- (27.5,-41.1);
\draw (27.5,-32.5) node [left] {$l$};
\draw [black] (19.6,-34.06) -- (53.8,-33.64);
\fill [black] (53.8,-33.64) -- (52.99,-33.15) -- (53.01,-34.15);
\draw [black] (44.5,-41.9) -- (44.5,-23.1);
\fill [black] (44.5,-23.1) -- (44,-23.9) -- (45,-23.9);
\draw (45,-32.5) node [right] {$m$};
\end{tikzpicture}
Each arrow represents a path, and (S,T) is the new edge being added.
\end{center}
\begin{theorem}
\label{reductionRule}
If conflicting paths $g_2 = f_2; (S,T); h_2$ then it must be that $g_1 = f_1; (S,T); h_1$.
\end{theorem} 
% monomorphisms
\begin{proof}
We use the fact that $f_1$=$l; f_2$ and $h_1$=$h_2; m$.
\[g_2 = f_2; (S,T); h_2 \Rightarrow l; g_2 = l; f_2; (S,T) ; h_2 \]
\[\Rightarrow l ; g_2 ; m = l ; f_2 ; (S,T) ; h_2 ; m \Rightarrow g_1 = f_1 ; (S,T) ; h_1\]
\end{proof}

Note that the proof is not affected if any of these paths is the identity, eg. if $f_1$ is the identity and S and $P_1$ are actually the same node.

@@

Using this information it is possible to choose a minimal subset of path pairs to verify.

@@

\begin{verbatim}
# Existng graph
Graph existingGraph;
# New edge
Edge (S, T);

predecessors = S.predecessors(existingGraph)
successors = T.successors(existingGraph)

# Construct the graph
Graph terminalPairGraph;
for q in successors:
    for p in predecessors:
        terminalPairGraph.addNode(q, p)
        # addEdge function should handle addition of edges between nodes that 
        # haven't been defined yet
        for q_pred in q.predecessors(existingGraph):
            for p_succ in p.successors(existingGraph):
                terminalPairGraph.addEdge((q_pred, p_succ))

# Now we reduce the graph.
verificationSet = {}
while (len(terminalPairGraph.nodes) > 0):
    # Choose a random node to start reduction from.
    current_node = terminalPairGraph.nodes[0]
    visited_nodes = set()
    # Climb up the graph till a "root" node is found
    while (len(currentNode.parents()) > 0):
        visited_nodes.add(currentNode)
        # Choose a parent at random.
        currentNode = currentNode.parents()[0]
        # Deal with cycles by reducing them
        if currentNode in visited_nodes:
            edges = getAllEdges(visited_nodes, terminalPairGraph)
            terminalPairGraph.removeNodes(visitedNodes)
            # All the incoming and outgoing edges from the entire visited set are now 
            # transferred to a single node that represents the class
            terminalPairGraph.addNode(currentNode, edges)
    # No predecessor! We should have reached a root!
    verificationSet.add(currentNode)
    terminalPairGraph.removeNodes(currentNode.successors(terminalPairGraph)

# This contains the minimal set of (source, sink) pairs whose verification 
# implies verification of the entire graph. 
return verificationSet

\end{verbatim}
\begin{algorithm}
    \caption{Optimal subset algorithm}
    \label{optimal}
    \KwData{placeholder}
    \KwResult{placeholder}
\end{algorithm}

The result of the algorithm cannot be further reduced. Also, the verification of the parallel pairs returned in the algorithm implies that the output of the previous algorithm must commute, and transitively with Theorem \ref{verifyingSet} that the entire diagram must commute.

@@

\subsection{Verification}
Ultimately verification is performed by calling an equality oracle (that is beyond the scope of this work) for every comparable pair returned by the previous stage.

\begin{verbatim}
for (path_1, path_2) in parallelPairs:
    if (path_1 != path_2):
        print($"Edge addition does not maintain path independence. 
        Counterexample: {path_1}, {path_2}")
        return False
print("Edge addition maintains path independence.")
return True
\end{verbatim}

\section{Applications}
TODO.
\subsection{Gator}
\subsection{Currency Graph}

@@

Using an API ~\footnote{https://exchangeratesapi.io}, we built the fully connected diagram of exchange rates between 32 currencies on a given day.
To ensure that it indeed commuted, we started with an empty diagram, and added in edges one by one.
Before the addition of each edge, we used the algorithms (~\ref{polynomial} and ~\ref{optimal}) to ensure the addition of a new edge did not introduce inconsistencies in the existing diagram.
If a new edge was, in fact, problematic, the algorithms returned an example inconsistent
pair that would arise from the addition of the edge.
The pair would consist of two currency transformation sequences with the 
same source currency and ultimate destination currency, but with different effective 
exchange rates values, as computed by taking the product of all the exchange rates
encountered through the chain.

We allowed an "error tolerance", so that differences reported would not be the trivial consequences of a floating point error.
However, this relaxation of the equality oracle into imprecision meant that the mathematical reasoning that allow the algorithms to remove redundant path checks no longer applied.
For instance, composing a new function with two approximately equal functions does not lead to equal results, so theorem ~\ref{reductionRule} fails with this approximate equality.
When the algorithms reported no inconsistencies, it was still possible that the graph possessed inconsistencies above the given threshold and did not commute.
Nonetheless, both algorithms were effective in catching inconsistencies. Algorithm ~\ref{polynomial} started finding inconsistencies at error tolerances to the order of $10^{-3}$, and ~\ref{optimal}, which makes more invalid redundant path check removals, at error tolerances to the order of $10^{-7}$.

Averaging over evaluation for the first 30 days of 2020, building and verifying a diagram to completion (inclusive of the time required by network calls) took 243$\pm$19 seconds using algorithm ~\ref{optimal}, and in 133$\pm$13 seconds with algorithm ~\ref{polynomial}. 
Computation was performed on a Macbook Pro 2015, 2.9 GHz Dual-Core Intel Core i5.
More information about implementation can be found in the evaluation section.

@@

Modified algorithm for diagrams without the identity:

\begin{verbatim}
Graph existingGraph;
Edge newEdge;

Set parallelPairs = new Set();
for S in existingGraph.Nodes():
    for T in existingGraph.Nodes():
        try:
            Path pathWithNewEdge = 
                FindPath(
                    graph: existingGraph, 
                    sourceNode: S,
                    sinkNode: newEdge.Source()) +
                newEdge as Path +
                FindPath(
                    graph: existingGraph, 
                    sourceNode: newEdge.Sink(), 
                    sinkNode: T)
            if (S==T):
                // Cycles are a special case.
                parallelPairs.add(pathWithNewEdge, pathWithNewEdge+pathWithNewEdge)
                for edge in (S.forwardEdges - pathWithNewEdge[0]):
                    parallelPairs.add(edge, edge+pathWithNewEdge)
                for edge in (S.backwardEdges - pathWithNewEdge[-1]):
                    parallelPairs.add(edge, pathWithNewEdge+edge)
            else:
                Path pathInOldGraph = FindPath(
                    graph: existingGraph, 
                    sourceNode: S, 
                    sinkNode: T);
                parallelPairs.add((pathInOldGraph, pathWithNewEdge));
        catch (pathFindingFailed Exception):
            // No comparable pairs from node {S} to node {T} 
            // that need to be checked
            continue
    cycle = findCycleWithTerminalNode(newEdge.Source()):
    if cycle:    
        parallelPairs.add(newEdge, cycle+newEdge)
    cycle = findCycleWithTerminalNode(newEdge.Source()):
    if cycle:    
        parallelPairs.add(newEdge, newEdge+cycle)
output parallelPairs;
\end{verbatim}

The optional minimization algorithm is modified to exclude nodes representing cycles where the source and sink are equal, since the reduction case no longer will apply in the absence of an implicit identity on the terminal node of the cycle.

>>>>

\labelformat{algocf}{algorithm\,(#1)}

@@

\begin{abstract}
Commutative diagrams arise in many systems, such as an implicit type conversion graph in some programming languages.
It can be useful in program analysis to verify that a diagram representing conversions remains commutative over the course of its construction.
A naive approach to solving the problem would take factorial time, and the presence of cycles could potentially lead to an infinite number of paths to check. Previous work provides an algorithm for the special case of acyclic diagrams that verifies that a given diagram commutes in $O(|V|^4|E|^2)$ time, where V is the set of vertices in the diagram, and E, the set of edges. 
We present an algorithm that given a diagram and an edge to add to it, verifies that the resultant diagram commutes in $O(|V|(|E| + |V|)$ time. For the case when checking the equality of paths is expensive, we present an optimization that runs in $O(V^4)$ time but reduces to the minimum possible number of equality checks. We evaluate the algorithms against baselines, and see its practical application in the compiler of a domain specific language for geometry types. As a demonstration of the more general applicability of the algorithm, we use it to identify discrepancies in a  currency conversion graph.
\end{abstract}

@@

We start with the set of all parallel pairs in the diagram (with the new edge added in), and pare it down to be finite by handling cycles.
We find all simple cycles in the diagram, using a procedure like Johnson's algorithm. 
\xxx[AK]{TODO cite the paper.}
We verify for each cycle that a single traversal starting on any node is equal to the identity function by adding pairs of (cycle starting at node, identity function) for each node in the function to the output verification set.
\xxx[AK]{Need cleaner way to put this.}
As cycles must now be the identity, for any pair in the set of all parallel paths, any instance of a cycle can be removed to obtain an equivalent pair with shorter, cycle free paths.
If the shorter pair has equal paths then the paths in the original pair must also be equal to each other.
It is therefore safe to remove all pairs of paths with cycles, leaving only parallel pairs where neither path has a cycle.
Since there are only a finite number of paths in the graph with no cycles, (to the order of n! as a path is an ordering on nodes), and the parallel pairs are a subset of the possibilities generated by choosing a first and second member from this finite set, the resultant set of all parallel pairs must be finite.
\xxx[AK]{TODO Check if O(n!n) = O(n!). What about $O((n!)^2)$?}
The output of the algorithm, then, is the union of the set of cycle verification pairs, and the parallel pairs found by doing an all-acyclic-path search from every node in the diagram with the new edge added.

@@

We present the idea diagrammatically in figure \ref{figure_two_flip}. Arrows represent path phases (note that these are the composition of many edges, not a single edge). The new edge is represented with a dashed arrow.

\begin{figure}
\begin{center}
\begin{tikzpicture}[scale=0.2]
\tikzstyle{every node}+=[inner sep=0pt]
\draw [black] (18.1,-36.1) circle (3);
\draw (18.1,-36.1) node {$S$};
\draw [black] (50.2,-36.1) circle (3);
\draw (50.2,-36.1) node {$T$};
\draw [black] (26,-22.9) circle (3);
\draw (26,-22.9) node {$F_1$};
\draw [black] (41.3,-22.9) circle (3);
\draw (41.3,-22.9) node {$F_2$};
\draw [dashed] (21.1,-36.1) -- (47.2,-36.1);
\fill [black] (47.2,-36.1) -- (46.4,-35.6) -- (46.4,-36.6);
\draw (34.15,-35.6) node [above] {new edge};
\draw [black] (24.46,-25.47) -- (19.64,-33.53);
\fill [black] (19.64,-33.53) -- (20.48,-33.1) -- (19.62,-32.58);
\draw [black] (29,-22.9) -- (38.3,-22.9);
\fill [black] (38.3,-22.9) -- (37.5,-22.4) -- (37.5,-23.4);
\draw [black] (48.52,-33.61) -- (42.98,-25.39);
\fill [black] (42.98,-25.39) -- (43.01,-26.33) -- (43.84,-25.77);
\end{tikzpicture}
\end{center}
\caption{Two flip tolerant path.}
\label{figure_two_flip}
\end{figure}

Here, (S, $F_1$, $F_2$, T) is a two flip tolerant path. ($F_1$, S, T, $F_2$) is a new path, created because of the addition of (S,T), that conflicts with ($F_1$, $F_2$).

@@

A diagram's structure can result in some redundancy, in the sense that verifying a subset of pairs can imply the verification of path pairs that aren't in the subset.
For example, in figure ~\ref{figure_redundancies}, verifying $p_1$ and $p_2$ implies that the paths in $p_3$ are equal too.

\begin{figure}

@@

$p_1$=($f_1;f_2$, $g_1$)\\
$p_2$=($j_1;j_2$, $h_1$)\\
$p_3$=($f_1;h_1$, $f_1;j_2$)
\end{center}
\caption{Path checking redundancies: verifying $p_1$ and $p_2$ implies that the paths in $p_3$ are equal.}
\label{figure_redundancies}
\end{figure}

The approach in this algorithm, at a high level, is to define a function that takes in a subset of pairs and returns the subset of pairs whose verification is implied by verifying the input set. Then, greedy elimination of redundancies is performed until a minimal set is reached.

@@

% A complete implementation of the described algorithm is included in the appendix.

Roughly, it proceeds by first efficiently finding a "spanning" set of bilinkings (a subset whose verification implies the verification of all bilinkings in the graph).
It does this by, starting at every node, finding the reachable subsection of the graph, and a spanning tree for the subsection.
From each edges in the reachable section that is not a part of the tree, it generates a bilinking using the edge and a path in the tree that is parallel to the edge (algorithm ~\ref{algo_spanning_set}).

\xxx[dg]{Probably put this in the appendix}
\begin{algorithm}
\label{alg_spanning_set}
\DontPrintSemicolon
\KwResult{Find a spanning set Rs = [$r_1$, ... ,$r_k$].}

Graph existingGraph\;
$R_s$ $\gets$ \{\}\;
\ForEach{node v in V}{
    S $\gets$ existingGraph.extractReachableSection(v)\;
    T $\gets$ createMinimumSpanningTree(S)\;
    excludedEdges = S.edges - T.edges\;
    \ForEach{edge e $\in$ excludedEdges}{
        firstPath = T.findPath(source: e.source, sink: e.sink)\;
        $R_s$.addElement(new Bilinking(firstPath, e)\;
    } 
}
\KwRet{$R_s$}\;

\caption{finding spanning set}\label{algo_spanning_set}
\end{algorithm}

With the spanning set thus initialized, it greedily tries to remove each pair from the spanning set if the set remains spanning even after removing the edge (algorithm ~\ref{algo_minimal_spanning_set}).

\begin{algorithm}
\DontPrintSemicolon
\KwResult{Find a minimal spanning set R.}

\SetKwFunction{FSigma}{$\sigma$}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\FSigma{inputSet S, codomain = Rs}}{
    output $\gets$ \{\}\;
    \For{bilinking $\in$ Rs}{
        smallerPairs $\gets$ \{\}\;
        \For{candidateBilinking $\in$ Rs}{
            \If{$\exists$ path from bilinking.source to candidateBilinking.source
            and
            $\exists$ path from candidateBilinking.sink to bilinking.sink}{
                \tcc{Check if there's a linear dependence between the bilinking and smallerPairs.
                    Vectorize(pair) returns a vector with a dimension for each edge in the graph,
                    with a 1 if pair includes that edge and 0 if it doesn't.}    
                matrix $\gets$ [vectorize(pair) for pair in smallerPair]\;
                matrix $\gets$ matrix+[vectorize(bilinking)]\;
                \If{determinant(matrix)\%2==0}{
                    output.add(bilinking)
                }
            }
        }
    }
    \KwRet{output}
}

R $\gets$ Rs\;
\For{i=1 to K}{
    \If{$r_i \in \sigma$(R-$r_i$)}{
       R $\gets$ R-$r_i$\;
    }
}
\KwRet{R}\;

\caption{Minimal spanning set}\label{algo_minimal_spanning_set}
\end{algorithm}

@@

We present a polynomial time solution to the \textsc{verification set} problem.
Like with the baseline, we do not concern ourselves with parallel pairs where neither or both paths pass through the new edge.
The key observation that allows us to improve on the online baseline is that, for a given source and sink pair, only a single parallel pair needs to be verified. 
This is an implication of Theorem \ref{reductionRule}, expanded on later.
Theorem \ref{verifyingSet} shows that should our selected set of pairs along with cycles passing through the new edge be verified commutative, the entire diagram must commute. 
The approach is to identify a parallel pair with exactly one path through the edge for each (source, sink) pair (algorithm ~\ref{algo_online_polynomial}).

\begin{algorithm}
\DontPrintSemicolon
\KwData{existing graph, new edge.}
\KwResult{Set of parallel pairs to verify.}
\SetKwProg{try}{try}{:}{}
\SetKwProg{catch}{catch}{:}{end}
Graph existingGraph;
Edge newEdge;

parallelPairs $\gets$ \{\}\;
\For{S in existingGraph.Nodes}{
    \For{T in existingGraph}{
        \try{}{
            Path pathWithNewEdge $\gets$ FindPath(
                graph: existingGraph, 
                sourceNode: S,
                sinkNode: newEdge.Source()) +\;
            newEdge as Path +\;
            FindPath(
                graph: existingGraph, 
                sourceNode: newEdge.Sink(), 
                sinkNode: T)\;
            \If{S == T}{
                pathInOldGraph $\gets$ identity(S)\;
            }
            \Else(){
                Path pathInOldGraph $\gets$ FindPath(
                    graph: existingGraph, 
                    sourceNode: S, 
                    sinkNode: T)\;
            }
            parallelPairs.add((pathInOldGraph, pathWithNewEdge))\;
        }
        \catch{PathFindingFailedException}{
           \tcc{No comparable pairs from node S to node T that need to be checked}
           continue\;
        }
    }
}
\KwRet{parallelPairs}\;
\caption{Online polynomial time algorithm to find parallel pair set}\label{algo_online_polynomial}
\end{algorithm}

@@

Consider the situation in figure ~\ref{figure_reduction_rule}.

\begin{figure}
\begin{center}
\begin{tikzpicture}[scale=0.2]
\tikzstyle{every node}+=[inner sep=0pt]
\draw [black] (16.6,-34.1) circle (3);
\draw (16.6,-34.1) node {$S$};
\draw [black] (56.8,-33.6) circle (3);
\draw (56.8,-33.6) node {$T$};
\draw [black] (28,-20.1) circle (3);
\draw (28,-20.1) node {$P_1$};
\draw [black] (44.5,-20.1) circle (3);
\draw (44.5,-20.1) node {$P_2$};
\draw [black] (28,-44.9) circle (3);
\draw (28,-44.9) node {$Q_1$};
\draw [black] (44.5,-44.9) circle (3);
\draw (44.5,-44.9) node {$Q_2$};
\draw [black] (31,-20.1) -- (41.5,-20.1);
\fill [black] (41.5,-20.1) -- (40.7,-19.6) -- (40.7,-20.6);
\draw (36.25,-20.6) node [below] {$g_1$};
\draw [black] (26.11,-22.43) -- (18.49,-31.77);
\fill [black] (18.49,-31.77) -- (19.39,-31.47) -- (18.61,-30.84);
\draw (21.74,-25.67) node [left] {$f_1$};
\draw [black] (25.82,-42.84) -- (18.78,-36.16);
\fill [black] (18.78,-36.16) -- (19.01,-37.08) -- (19.7,-36.35);
\draw (23.57,-39.02) node [above] {$f_2$};
\draw [black] (31,-44.9) -- (41.5,-44.9);
\fill [black] (41.5,-44.9) -- (40.7,-44.4) -- (40.7,-45.4);
\draw (36.25,-45.4) node [below] {$g_2$};
\draw [black] (54.59,-35.63) -- (46.71,-42.87);
\fill [black] (46.71,-42.87) -- (47.64,-42.7) -- (46.96,-41.96);
\draw (49.21,-38.76) node [above] {$h_2$};
\draw [black] (54.78,-31.38) -- (46.52,-22.32);
\fill [black] (46.52,-22.32) -- (46.69,-23.25) -- (47.43,-22.57);
\draw (51.19,-25.39) node [right] {$h_1$};
\draw [black] (28,-23.1) -- (28,-41.9);
\fill [black] (28,-41.9) -- (28.5,-41.1) -- (27.5,-41.1);
\draw (27.5,-32.5) node [left] {$l$};
\draw [black] (19.6,-34.06) -- (53.8,-33.64);
\fill [black] (53.8,-33.64) -- (52.99,-33.15) -- (53.01,-34.15);
\draw [black] (44.5,-41.9) -- (44.5,-23.1);
\fill [black] (44.5,-23.1) -- (44,-23.9) -- (45,-23.9);
\draw (45,-32.5) node [right] {$m$};
\end{tikzpicture}
Each arrow represents a path, and (S,T) is the new edge being added.
\end{center}
\caption{Reduction rule.}
\label{figure_reduction_rule}
\end{figure}

\begin{theorem}
\label{reductionRule}
If conflicting paths $g_2 = f_2; (S,T); h_2$ then it must be that $g_1 = f_1; (S,T); h_1$.
\end{theorem} 
% monomorphisms
\begin{proof}
We use the fact that $f_1$=$l; f_2$ and $h_1$=$h_2; m$.
\[g_2 = f_2; (S,T); h_2 \Rightarrow l; g_2 = l; f_2; (S,T) ; h_2 \]
\[\Rightarrow l ; g_2 ; m = l ; f_2 ; (S,T) ; h_2 ; m \Rightarrow g_1 = f_1 ; (S,T) ; h_1\]
\end{proof}

The proof is not affected if any of these paths is the identity, eg. if $f_1$ is the identity and S and $P_1$ are actually the same node.

@@

Using this information it is possible to choose a minimal subset of path pairs to verify, as in algorithm ~\ref{algo_online_minimal}.

@@

\begin{algorithm}
\DontPrintSemicolon
\KwData{Existing graph, new edge.}
\KwResult{Set of parallel pairs to verify.}
Graph existingGraph\;
Edge (S, T)\;

predecessors $\gets$ S.predecessors(existingGraph)\;
successors $\gets$ T.successors(existingGraph)\;

Graph terminalPairGraph\;
\For{q $\in$ successors}{
    \For{p $\in$ predecessors}{
        terminalPairGraph.addNode(q, p)\;
        \For{qPred $\in$ q.predecessors(existingGraph)}{
            \For{pSucc $\in$ p.successors(existingGraph)}{
                terminalPairGraph.addEdge((qPred, pSucc))\;
            }
        }
    }
}

verificationSet $\gets$ \{\}\;
\While{len(terminalPairGraph.nodes) $>$ 0}{
    currentNode $\gets$ terminalPairGraph.nodes[0]\;
    visitedNodes $\gets$ \{\}\;
    \While{len(currentNode.parents()) $>$ 0}{
        visitedNodes.add(currentNode)\;
        currentNode $\gets$ currentNode.parents()[0]\;
        \If{currentNode $\in$ visitedNodes}{
            edges $\gets$ getAllEdges(visitedNodes, terminalPairGraph)\;
            terminalPairGraph.removeNodes(visitedNodes)\;
            terminalPairGraph.addNode(currentNode, edges)\; 
        }
    }
    verificationSet.add(currentNode)\;
    terminalPairGraph.removeNodes( currentNode.successors(terminalPairGraph))\;
}
\KwRet{verificationSet}\;
\caption{Minimal set finding algorithm}\label{algo_online_minimal}
\end{algorithm}

The result of the algorithm cannot be further reduced.
Also, the verification of the parallel pairs returned in the algorithm implies that the output of the previous algorithm must commute, and transitively with Theorem ~\ref{verifyingSet} that the entire diagram must commute.

@@

\subsection{Verification}
Ultimately verification is performed by calling an equality oracle (that is beyond the scope of this work) for every comparable pair returned by the previous stage, as in algorithm ~\ref{algo_verification}.

\begin{algorithm}
\DontPrintSemicolon
\For{(path1, path2) $\in$ parallelPairs}{
    \If{path1 $!$= path2}{
        \KwRet{False}\;
    }
}
\KwRet{True}\;
\caption{Verification algorithm}\label{algo_verification}
\end{algorithm}

\section{Applications}
TODO.

\subsection{Gator}

\subsection{Currency Graph}

@@

Using an API ~\footnote{https://exchangeratesapi.io}, we built the fully connected diagram of exchange rates between 32 currencies on a given day.
To ensure that it indeed commuted, we started with an empty diagram, and added in edges one by one.
Before the addition of each edge, we used the algorithms (~\ref{algo_online_polynomial} and ~\ref{algo_online_minimal}) to ensure the addition of a new edge did not introduce inconsistencies in the existing diagram.
If a new edge was, in fact, problematic, the algorithms returned an example inconsistent
pair that would arise from the addition of the edge.
The pair would consist of two currency transformation sequences with the 
same source currency and ultimate destination currency, but with different effective 
exchange rates values, as computed by taking the product of all the exchange rates
encountered through the chain.

We allowed an "error tolerance", so that differences reported would not be the trivial consequences of a floating point error.
However, this relaxation of the equality oracle into imprecision meant that the mathematical reasoning that allow the algorithms to remove redundant path checks no longer applied.
For instance, composing a new function with two approximately equal functions does not lead to equal results, so theorem ~\ref{reductionRule} fails with this approximate equality.
When the algorithms reported no inconsistencies, it was still possible that the graph possessed inconsistencies above the given threshold and did not commute.
Nonetheless, both algorithms were effective in catching inconsistencies. Algorithm ~\ref{algo_online_polynomial} started finding inconsistencies at error tolerances to the order of $10^{-3}$, and ~\ref{algo_online_minimal}, which makes more invalid redundant path check removals, at error tolerances to the order of $10^{-7}$.

Averaging over evaluation for the first 30 days of 2020, building and verifying a diagram to completion (inclusive of the time required by network calls) took 243$\pm$19 seconds using algorithm ~\ref{algo_online_minimal}, and in 133$\pm$13 seconds with algorithm ~\ref{algo_online_polynomial}. 
Computation was performed on a Macbook Pro 2015, 2.9 GHz Dual-Core Intel Core i5.
More information about implementation can be found in the evaluation section.

@@
